---
title: "Sphingolipid_PCA_Biplot_Heatmap_Analysis"
author: "Napier Lab"
date: "2023-04-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load packages
```{r}
require(RColorBrewer)
require(ggplot2)
require(ggfortify)
require(pheatmap)
require(reshape2)
require(gplots)
require(iheatmapr)
require(plyr)
require(ggdendro)
require(glmulti)
require(factoextra)
require(readr)
require(ggplot)
require(rJava)
require(dplyr)
require(tibble)
require(stringr)
require(FactoMineR)
```

#Import the data, scale the lipidomic data, add back in diet group names
```{r}
setwd("~/Desktop/Amy Fatty Acid Data/PC Data And Analysis") #change file path to match the location of your data
NIH_PC_SCKD <- read.csv("PCsSCKD.csv", header= T, blank.lines.skip = TRUE, as.is=)
rownames(NIH_PC_SCKD) <- as.character(NIH_PC_SCKD$X)
NIH_PC_SCKD$X <- NULL
NIH_PC_SCKD$names <- rownames(NIH_PC_SCKD)
head(NIH_PC_SCKD)

pca_NIH_PC_SCKD <- scale(NIH_PC_SCKD[1:65]) #select only the quantitative lipidomic data for scaling
pca_NIH_PC_SCKD[is.na(pca_NIH_PC_SCKD)] <- 0 # set NA values to O
pca_NIH_PC_SCKD <- as.data.frame(pca_NIH_PC_SCKD)
pca_NIH_PC_SCKD$diet <- NIH_PC_SCKD$Diet #assign diet groups to scaled data
pca_NIH_PC_SCKD$names <- NIH_PC_SCKD$names
head(pca_NIH_PC_SCKD)
```

#Creation of PCA with confidence ellipses (Figure 1A)
```{r}
groups <- as.factor(pca_NIH_PC_SCKD$diet) #assign sample diet groups as a factor
groups <- droplevels(groups)
fviz_pca_ind(prcomp(pca_NIH_PC_SCKD[1:65]),
             col.ind = groups, # color by groups
             palette = c('#FC8008', '#807F7F'),
             addEllipses = TRUE, # Concentration ellipses
             ellipse.type = "confidence", #set to confidence ellipse
             ellipse.level=0.95, #settting for what % confidence
             legend.title = "Diet",
             repel = TRUE,
             labels=TRUE,
             labelsize=8,
             title="",
             show.legend=TRUE,
              mean.point=FALSE,
             label="none",
            legend= "none",
             pointsize=5,
             pointshape=19,
            
) +
  theme(text = element_text(family="Arial", size = 24), #all of ths is for changing the look of the image
        axis.title = element_text(family="Arial", size = 24),
        axis.text = element_text(family="Arial", size = 24), 
        legend.text= element_text(family="Arial", size=24), 
        #legend.position= "right",
        panel.grid.major=element_blank(), 
        panel.grid.minor=element_blank(), 
        axis.line=element_line()
) +
  xlim(-12,12) + ylim(-5,5)

```

#Creation of the sphingolipid biplot. In this context LS1, LS2, and LS3 in the resulting figure are the same as SC1, SC2, and SC3. Sphingolipid variables (vectors) that are close together and form small angles with one another are positively correlated. Variables point in the direction of the principal component(s) they strongly influence. In the sphingolipid biplot, all of the variables point along the PC1 axis towards the KD samples indicated that the KD samples are relatively more enriched in these spingolipids. This enrichment strongly contributes to the separation of the KD and SC as the SC samples group towards the opposite side of PC1, away from where the variables are pointing.
```{r}

fviz_pca_biplot(prcomp(pca_NIH_PC_SCKD[1:65]),
             select.var =  list(contrib=5),
             repel = TRUE,
             col.ind = groups, # color by groups
             palette = c('#FC8008', '#807F7F'),
             labelsize=5,
             title="",
             show.legend=TRUE,
              mean.point=FALSE,
             labels=TRUE,
            legend= TRUE,
            legend.title= "Diet",
             pointsize=4,
             pointshape=19,
            #axes.linetype=NA
) +
  theme(text = element_text(family="Arial", size = 24),
        axis.text = element_text(family="Arial", size = 24), 
        legend.text= element_text(family="Arial", size=24), 
        legend.position= "right",
        panel.grid.major=element_blank(), 
        panel.grid.minor=element_blank(), 
        axis.line=element_line()
) 

```

#Creation of sphingolipid heatmap (Figure 1C)
```{r}
data <- read.csv("/Users/hickman6/Desktop/Amy Fatty Acid Data/PC Data And Analysis/pHeatmapPCsSCKD.csv", header= T, blank.lines.skip = TRUE, as.is=) #change file path to match where your data file is located. This file has the data transposed to work with the pheatmap function and the group and time data removed. Instead of having a separate file you can also use the t() function to transpose the data and specifically choose the fatty acid data.
rownames(data) <- as.character(data$X)
data$X <- NULL
data <- as.data.frame(data)

my_sample_col <- data.frame(Diet = factor(rep(c("KD", "SC"),each=3))) #assigning diet groups to samples
row.names(my_sample_col) <- colnames(data)
my_sample_col$Diet = factor(my_sample_col$Diet, levels = c("KD", "SC")) #create factor with diets
my_color = list(Diet = c(KD="#FC8008", SC="#807F7F")) #assign colors to diets for the heatmap

pheatmap(data, 
         scale="row", #scale the data based on rows
         fontsize_row= 6, 
         color=colorRampPalette(c("blue", "white", "red"))(50), 
         cellwidth=8, 
         cellheight=8,
         treeheight_row=0, 
         treeheight_col=12, 
         annotation_col = my_sample_col, 
         annotation_colors =my_color,
         legend=TRUE)
```
